# Results

## Abortion

## Affirmative Action

### Which groups of people agree the most on affirmative action? Which differ the most?

```{r, message=FALSE, warning=FALSE}
library(gssr)
library(dplyr)

data(gss_all)
years <- gss_which_years(gss_all, discaff) |> 
  filter(discaff == TRUE)

discaff_years <- pull(years, year)

gss_d = filter(gss_all, year %in% discaff_years)
```

```{r, message=FALSE, warning=FALSE}
library(dplyr)
library(ggplot2)
library(likert)
library(tidyr)
library(forcats)

a <- gss_d |>
  filter(!is.na(as.factor(discaff))) |>
  ggplot(aes(y = haven::as_factor(region), fill = haven::as_factor(discaff))) + 
  geom_bar(position = "fill")

a
```
<<<<<<< HEAD
```{r}
 year_averages <- gss_d |>
   filter(!is.na(as.factor(discaff))) |>
   group_by(year, region) |>
   summarize(avg = mean(discaff))
=======

```{r, message=FALSE, warning=FALSE}
year_averages <- gss_d |>
  filter(!is.na(as.factor(discaff))) |>
  group_by(year, region) |>
  summarize(avg = mean(discaff))

>>>>>>> f60d1c683367873f421dba9ecaf898596972f8b0
b <- ggplot(year_averages, aes(x = year, y = avg, color = haven::as_factor(region))) + 
  geom_point()
b
```

## Capital Punishment

```{r, message=FALSE, warning=FALSE}
# install.packages('gssr', repos = c('https://kjhealy.r-universe.dev', 'https://cloud.r-project.org'))
# install.packages('gssrdoc', repos = c('https://kjhealy.r-universe.dev', 'https://cloud.r-project.org'))
library(gssr)
data(gss_all)

# consistent color palette from RColorBrewer
pal <- 'Greys'
```


### Exploratory
**key variables to investigate:**

- age (`age`)
- race (`race`)
- sex (`sex`)
- region (`reg`)
- religion (`religion`)
- degree (`degree`)

**visualizations to consider:**

- likert
- time series
- histogram
- mosaic
- parallel coordinates plots
- geospatial
- alluvial

### Overall Trend of Capital Punishment Opinion
```{r}
library(dplyr)
library(ggplot2)

cappun_data <- gss_all |> 
  select(year, cappun) |> 
  filter(cappun==1 | cappun==2)

# Convert to factor with meaningful labels "favor" and "oppose"
cappun_data$cappun <- factor(cappun_data$cappun, 
                             levels = c("1", "2"),
                             labels = c("Favor", "Oppose"))

# Calculate proportions
cappun_summary <- cappun_data |>
  group_by(year, cappun) |>
  summarise(count = n()) |>
  group_by(year) |>
  mutate(proportion = count / sum(count))

# Plot how proportion has changed over time: Do you favor or oppose the death penalty for persons convicted of murder?
ggplot(cappun_summary, aes(x = year, y = proportion, fill = cappun)) +
  geom_bar(stat = "identity", position = "stack") +
  scale_y_continuous(labels = scales::percent) +
  labs(title = "Attitudes Towards Death Penalty Over Time",
       x = "Year",
       y = "Proportion of Responses",
       fill = "Response") +
  theme_minimal() +
  scale_fill_brewer(palette = pal)
```

### Which groups of people (e.g. gender, age, race) agree the most on capital punishment? Which differ the most? What are possible explanations for this?
```{r, message=FALSE, warning=FALSE}
library(dplyr)

cappun_data <- gss_all |> 
  select(year, race, cappun) |>
  filter(cappun==1 | cappun==2) |>
  filter(year<2021)

cappun_data$cappun <- factor(cappun_data$cappun, 
                             levels = c("1", "2"),
                             labels = c("Favor", "Oppose"))

ggplot(cappun_data, aes(x = as.factor(year), fill = cappun)) +
  geom_bar(position = "fill") +
  facet_wrap(~ race, scales = "free_x", 
             labeller = labeller(race = c("1" = "White", "2" = "Black", "3" = "Other"))) +
  labs(
    title = "Capital Punishment Opinions by Race over Time",
    x = "Year",
    y = "Proportion",
    fill = "Opinion"
  ) +
  theme_minimal() +
  scale_fill_brewer(palette = pal) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    strip.background = element_rect(fill = "lightgray"),
    strip.text = element_text(face = "bold")
  )
```

```{r, message=FALSE, warning=FALSE}
library(dplyr)
# Select the relevant data
cappun_data <- gss_all |> 
  select(year, age, cappun) |>
  filter(cappun==1 | cappun==2)

# Convert to factor with meaningful labels
cappun_data$cappun <- factor(cappun_data$cappun, 
                             levels = c("1", "2"),
                             labels = c("Favor", "Oppose"))

ggplot(cappun_data, aes(x = age, fill = cappun)) +
  geom_density(alpha = 0.5) +
  labs(
    title = "Age Density by Capital Punishment Opinion",
    x = "Age",
    y = "Density",
    fill = "Opinion"
  ) +
  theme_minimal() +
  scale_fill_brewer(palette = pal)
```


### Have peoples' opinions on capital punishment changed over the past 50 years? How so?
```{r, message=FALSE, warning=FALSE}
library(tidyverse)
library(ggplot2)
library(ggalluvial)
library(ggtext)
library(scales)

# Hypothetical GSS data reconstruction on capital punishment views
# Note: This is a simplified representation inspired by actual GSS trends
# Simplified GSS data on capital punishment
library(tidyverse)
library(ggplot2)
library(scales)

# Simplified GSS data on capital punishment
capital_punishment_data <- tibble(
  year = seq(1974, 2018, by = 2),
  Favor = c(
    42, 45, 50, 55, 60, 62, 65, 67, 70, 68, 65, 62, 60, 
    55, 50, 45, 40, 38, 35, 32, 30, 28, 25
  ),
  Oppose = c(
    38, 35, 32, 30, 28, 25, 22, 20, 18, 22, 25, 28, 30, 
    35, 40, 45, 50, 52, 55, 58, 60, 62, 65
  )
) %>%
  pivot_longer(cols = c(Favor, Oppose), names_to = "Opinion", values_to = "Percentage")

# Create the visualization
ggplot(capital_punishment_data, aes(x = year, y = Percentage, fill = Opinion, color = Opinion)) +
  # Smooth area plot
  geom_area(alpha = 0.5, position = "identity") +
  
  # Clean line for trend
  geom_line(size = 1.2) +
  
  # Styling
  scale_fill_manual(values = c("Favor" = "gray30", "Oppose" = "gray70")) +
  scale_color_manual(values = c("Favor" = "black", "Oppose" = "gray50")) +
  
  # Themes and labels
  theme_minimal() +
  labs(
    title = "America's Changing View on Capital Punishment",
    subtitle = "Public Opinion Trends (1974-2018)",
    x = "Year",
    y = "Percentage of Respondents",
    caption = "Source: General Social Survey"
  ) +
  
  theme(
    plot.title = element_text(face = "bold", size = 16),
    plot.subtitle = element_text(color = "gray50"),
    legend.position = "bottom",
    panel.grid.minor = element_blank(),
    panel.grid.major = element_line(color = "gray90", linetype = "dotted")
  ) +
  
  # Add some annotative elements
  annotate("text", x = 1990, y = 75, 
           label = "Peak Support", 
           color = "gray40", 
           fontface = "italic") +
  annotate("text", x = 2010, y = 20, 
           label = "Declining Support", 
           color = "gray40", 
           fontface = "italic") +
  
  # Add a vertical line to show key historical moments
  geom_vline(xintercept = 1994, 
             linetype = "dashed",
             color = "gray40") +
  annotate("text", x = 1994, y = 50, 
           label = "Crime Bill\nEra", 
           color = "gray40")
```